Refinement for open automata

### 2021-06-10 2021-Refinement/report.tex
Fixed many quick points

% transitions between sections
% TODO: CITER LES CONFERENCES

@@

We begin with Section \ref{sec:notations} by giving some notations used throughout this paper.
Then there are several sections dedicated to define all the interesting objects and properties:
In Section \ref{sec:def} we give a clean and up to date definition of open automata.
In Section \ref{sec:comp} we define their composition without requiring the use of open pNets.
In Section \ref{sec:proofelts} we adapt the standard properties related to refinement relation from LTS to open automata.
After that Section \ref{sec:prelref} is dedicated to progressively build a refinement relation.
Section \ref{sec:refinement} introduces and analyses the refinement relation on open automata.
% flat model
% future work
% conclusion

@@

The important properties we consider for the relation on states are being a simulation and the prevention of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
This definition translated to open automata the classical definition of simulation \marginpar{cite: Fernandez-Mounier-91-b}.
The last formula means that for every pair of related configurations of the automaton and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update. % OR variableS
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

We restrict the holes of the related automata so that we can explain the different refinements without their interactions.
In the following definition, \(\triangle \in \mbrc{\subseteq, =, \supseteq}\).
\begin{defi}[Hole-\(\triangle\) simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\triangle\) simulation of \(A_1\) by \(A_2\) if:
\defitem \(J_1 \triangle J_2\)
\defitem \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\defitem \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
	\mpar{\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} \defobject \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\
		\quad \mpar{\forall x \in X, J'_{2x} \cap J_1 = J'_1 \cap J_2} \\[-10pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap J_1} \beta_{1j} = \beta_{2xj} \\[12pt]
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}
	\end{array}} \\
	\wedge V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}
\end{multline*}
\end{defi}
The third item of the definition has two parts:
The first part is the main requirement, a (not equivalent) reordering of the pre-simulation requirement so that it better fits SMT.
Instead of matching an effective transition to another effective transition as in the pre-simulation definition, we match an open transition to a family of covering open-transitions.
The second part is directly the deadlock prevention requirement.

% TODO: expand on first part, especially the impact of /_\ on the different refinement allowed and what it means, and the fact that = could be substituted with \subseteq.

Examples of application of the different refinement relations are given in appendix \ref{apx:refrel}.
\begin{thm} A hole-\(\triangle\) simulation is a pre-simulation. \end{thm}
The condition \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\) is in both definitions so proving the following is sufficient:
We gather the existential variables in a valuation:
	\nu: \biguplus_{x \in X} \fvars{t_{2x}} \to \values \\
	\sigma \uplus \nu \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \bigvee_{x \in X} \mpar{\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x}} \beta_{1j} = \beta_{2xj} \wedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}} \hyp{4}
The left side of the implication is proven by \hyp{1}, and we destruct the disjuntion on the right side to get \(x \in X\) and the hypothesis
\[ \sigma \uplus \nu \vdash \alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap J_1} \beta_{1j} = \beta_{2xj} \wedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}} \hyp{4'} \]
Hence \(J'_{2x} \cap J_1 = J'_{2x} \cap J'_1\) in \hyp{4'}.
We can get back to prove \goal{0'} with the transition \(t_{2x}\) and the valuation \(\nu\) by \hyp{4'}.

\begin{thm} The hole-\(\triangle\) simulation on automata is a preorder. \end{thm}
Reflexivity is immediate with \(R \defobject \mpar{s, s'} \mapsto \choice{\bigwedge_{v \in V} v_1 = v_2 & s = s'\\ \bot & s \neq s'}\) where \(v_x\) is the renaming to distinguish conflicting automaton variables.

For three open automata \(A_1 \defobject \OAg[1]\), \(A_2 \defobject \OAg[2]\) and \(A_3 \defobject \OAg[3]\), such that \(R_{12}\) is a hole-\(\triangle\) simulation of \(A_1\) by \(A_2\) and \(R_{23}\) of \(A_2\) by \(A_3\), we want to define \(R_{13}\) by saying that there exists a state \(s_2 \in S_2\) and values for \(V_2\) such that \(R_{12}\mpar{s_1, s_2} \wedge R_{23}\mpar{s_2, s_3}\).
However we cannot talk about states in formulas.
The solution is to expand the existential quantifier on states as we did in deadlock prevention, which is possible because states are finite.
\nmm{R_{13} \defobject \mpar{s_1, s_3} \mapsto \exists V_2, \bigvee_\subbox{s_2 \in S_2} R_{12}\mpar{s_1, s_2} \wedge R_{23}\mpar{s_2, s_3}} is a witness of the transitivity, proof is given in appendix \ref{apx:trans}.
\begin{thm}[Context refinement] Hole-equal simulation is context refining \end{thm}
	For three open automata \(A_1 \defobject \OAg[1]\), \(A_2 \defobject \OAg[2]\) and \(A_3 \defobject \OAg[3]\), let \(k \in J_1\), \(A_{13} \defobject A_1\subst{A_3}{k}\) and \(A_{23} \defobject A_2\subst{A_3}{k}\) with \(A_{13} = \OAg[13]\) and \(A_{23} = \OAg[23]\).
	The idea for the relation on configurations is that \(A_2\) simulates \(A_1\) while \(A_3\) does the same things in both compositions.
	We pose
	\[ R \defobject \mpar{\mpar{s_1, s_{31}}, \mpar{s_2, s_{32}}} \mapsto \choice{R_{12}\mpar{s_1, s_2} \wedge \reach{A_{13}}\mpar{s_1, s_{31}} \wedge \bigwedge_{v_3 \in V_3} v_{31} = v_{32} & \text{if } s_{31} = s_{32} \\ \bot & \text{otherwise}} \]
	We want to prove that \(R\) is a hole-equal simulation of \(A_{12}\) by \(A_{23}\):
		\iff & \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R_{12}\mpar{s_{01}, s_{02}}} \wedge \mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}} \wedge \bigwedge_{v_3 \in V_3} \sigma_{031}\mpar{v_{31}} = \sigma_{032}\mpar{v_{32}} \\
		\end{array}\!\!} \goal{1}
	\[ V_{13} \uplus V_{23} \uplus \biguplus_\subbox{t_{23} \in \fOT{s_{23}}} \fvars{t_{23}} \vdash R\mpar{s_{13}, s_{23}} \wedge \bigvee_\subbox{t_{23} \in \fOT{s_{23}}} \fguard{t_{23}} \implies \bigvee_\subbox{t_{13} \in \fOT{s_{13}}} \fguard{t_{13}} \goal{2} \]
		\end{array}} \hyp{5}
	Transition \(t_{13}\) is obtained in \(A_{13}\) by composition of two transitions \nmm{t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}} and \nmm{t_3 \defobject \OTx{3}{}{3}{3} \in \fOT{s_3}} if \(k \in J'_1\) or by a the first transition and a fixed state \(s_3 \in S_3\) if \(k \notin J'_1\).
	We use \hyp{5} with the transition \(t_1\) to get
		\forall x \in X, J'_{2x} \cap J_1 = J'_1 \cap J_2 \hyp{5'} \\
		\end{array}} \hyp{5"}
		& = J'_{2x} \cap J_1 & \text{by \hyp{5'}} \\
		& = J'_{2x} \hyp{6}
	Using \hyp{6} we can build the familly of transition generated by the family \(t_{2x}^{x \in X}\):
		\OT{\mpar{s_2, s_3}}{\mpar{s'_{2x}, s'_3}}{\alpha_{2x}}{\beta_{2xj}^{j \in J'_{2x} \setminus \mbrc{k}} \uplus \beta_{3j}^{j \in J'_3}}{g_{2x} \wedge g_3 \wedge \alpha_3 = \beta_{2xk}}{\psi_{2x} \uplus \psi_3} & \text{if } k \in J'_{2x} \\
		\OT{\mpar{s_2, s_3}}{\mpar{s'_{2x}, s_3}}{\alpha_{2x}}{\beta_{2xj}^{j \in J'_2}}{g_2}{\psi_2} & \text{if } k \notin J'_{2x}
	} \]
	We will prove \goal{1} for the family \(t_{23x}^{x \in X}\), the new goals are
		\forall x \in X, J'_{23x} \cap J_{13} = J'_{13} \cap J_{23} \goal{1'} \\
		\end{array}} \goal{3}
\item[\goal{1'}:] Let \(x \in X\), by case on \(k \in J'_1\):
		\text{If } k \in J'_1: && J'_{23x} \cap J_{13} & = \mpar{J'_3 \cup J'_{2x} \setminus \mbrc{k}} \cap \mpar{J_3 \cup J_1 \setminus \mbrc{k}} \\
		&&& = \mpar{J'_3 \cap J_3} \cup \mpar{J'_{2x} \cap J_1} \setminus \mbrc{k} \\
		&&& = \mpar{J'_3 \cap J_3} \cup \mpar{J'_1 \cap J_2} \setminus \mbrc{k} & \text{by \hyp{5'}} \\
		&&& = \mpar{J'_3 \cup J'_1 \setminus \mbrc{k}} \cap \mpar{J_3 \cup J_2 \setminus \mbrc{k}} \\
		&&& = J'_{2x} \cap J_1 & \text{by } k \notin J'_1 = J'_{2x} \text{ \hyp{6}} \\
		&&& = J'_1 \cap J_2 & \text{by \hyp{5'}} \\
\item[\goal{3}:] Let \(\sigma: V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \to \values\).
	\[ \sigma \vdash R\mpar{s_{13}, s_{23}} \wedge g_{13} \hyp{7} \]
	Hypothesis immediately breaken down into
		\sigma \vdash R_{12}\mpar{s_1, s_2} \hyp{7a} \\
		\sigma \vdash \reach{A_{13}}\mpar{s_{13}} \hyp{7b} \\
		\sigma \vdash \bigwedge_{v_3 \in V_3} v_{31} = v_{32} \hyp{7c} \\
		\sigma \vdash g_{13} \hyp{7d}
	We use \hyp{5"} with valuation \(\sigma\) on variables \(V_1 \uplus V_2 \uplus \fvars{t_1}\), which is possible because the composition of transitions doesn't remove any variable (the simplification steps do, they give bisimilar automata).
	The left side of the implication is proved with \hyp{7a} and \hyp{7d}.
		\sigma \uplus \nu \vdash \alpha_1 = \alpha_{2x} \hyp{8a} \\
		\sigma \uplus \nu \vdash \bigwedge_\subbox{j \in J'_{2x} \cap J_1} \beta_{1j} = \beta_{2xj} \hyp{8b} \\
		\sigma \uplus \nu \vdash g_{2x} \hyp{8c} \\
		\sigma \uplus \nu \vdash R_{12}\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}} \hyp{8d}
	If \(k \in J'_1\), we extend the valuation \(\nu\) to cover the variables of \(t_3\) in \(t_{23x}\) using their value in \(t_{13}\), \(\nu' \defobject \nu \uplus \mset{v_{32} \mapsto \sigma\mpar{v_{31}}}{v \in \fvars{t_3}}\)
	We progress on \goal{3} by giving the valuation \(\sigma \uplus \nu'\) to the implicit \(\exists\), admitting the left side of the implication and choosing to prove the branch \(x\) of the disjunction:
	Because composition doesn't change the produced actions nor its variables, \goal{3a} is proved by \hyp{8a}.
	\goal{3b} holds because the valuations for the actions of the holes in \(J_3\) coincide by definition of \(\nu'\) and by \hyp{8b} for the others.
	For \goal{3c} we have two cases, if \(k \notin J'_1\) then \(g_{23x} \equiv g_{2x}\) and \hyp{8c} is sufficient, if \(k \in J'_1\) then \(g_{23x} \equiv g_{2x} \wedge g_3 \wedge \alpha_3 = \beta_{2xk}\) so we also need \hyp{7d} (\(g_{13} \equiv g_1 \wedge g_3 \wedge \alpha_3 = \beta_{1k}\)) and \hyp{8b} for the value \(k\) (\(\beta_{1k} = \beta_{2xk}\)).
	\goal{3d} is proved using \hyp{8d}, \hyp{7b} with the preservation of reachability across transitions (+ \hyp{7d}), \hyp{7c} and the fact that \(A_3\) performs the same transition with same values in both composed automata:
		\iff & \sigma \uplus \nu' \vdash \mpar{R_{12}\mpar{s'_1, s'_{2x}} \wedge \reach{A_{13}}\mpar{s'_{13}} \wedge \bigwedge_{v_3 \in V_3} v_{31} = v_{32}}\psubst{\psi_1 \uplus \psi_{31} \uplus \psi_{2x} \uplus \psi_{32}} \\
		\iff & \sigma \uplus \nu' \vdash R_{12}\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}} \wedge \reach{A_{13}}\mpar{s'_{13}}\psubst{\psi_{13}} \wedge \bigwedge_{v_3 \in V_3} \psi_{31}\mpar{v_{31}} = \psi_{32}\mpar{v_{32}} \\
		\iff & \mpar{\sigma \uplus \nu \vdash R_{12}\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}} \wedge \mpar{\sigma \vdash \reach{A_{13}}\mpar{s'_{13}}\psi_{13}} \wedge \mpar{\sigma \vdash \bigwedge_{v_3 \in V_3} v_{31} = v_{32}}
\item[\goal{2}:] Let \(\sigma: V_{13} \plus V_{23} \uplus \biguplus_\subbox{t_{23} \in \fOT{s_{23}}} \fvars{t_{23}} \to \values}.
		\sigma \vdash R_{12}\mpar{s_1, s_2} \hyp{9a} \\
		\sigma \vdash \reach{A_{13}}\mpar{s_{13}} \hyp{9b} \\
		\sigma \vdash \bigvee_\subbox{t_{23} \in \fOT{s_{23}}} \fguard{t_{23}} \hyp{9c}
		V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R_{12}\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \hyp{10} \\
		V_{13} \uplus \biguplus_\subbox{t_1 \in \fOT{s_1}} \fvars{t_1} \vdash \reach{A_{13}}\mpar{s} \wedge \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \implies \bigvee_\subbox{t_{13} \in \fOT{s_{13}}} \fguard{t_{13}} \hyp{11}
	We use \hyp{10} with valuation \(\sigma\) projected on the variables from generating transitions in \(A_2\) to get a valuation \nmm{\nu: \biguplus_\subbox{t_1 \in \fOT{s_1}} \fvars{t_1}}.
	The left side of the implication is proved using \hyp{9a} and \hyp{9c} because the guards from transitions of \(A_2\) are still in transitions of \(A_{23}\), the right side gives:
	\[ \sigma \uplus \nu \vdash \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \hyp{10'} \]
	We then use \hyp{11} with valuation \(\sigma \uplus \nu\) to get \nmm{\nu': \biguplus_\subbox{t_{13} \in \fOT{s_{13}}} \fvars{t_13}}.
	The left side of the implication is proved using \hyp{9b} and \hyp{10'}, the right side gives:
	\[ \sigma \uplus \nu' \vdash \bigvee_\subbox{t_{13} \in \fOT{s_{13}}} \fguard{t_{13}} \hyp{11'} \]
	\goal{2} is proved for valuation \(\nu'\) by admitting the left side and using \hyp{11'}.
\begin{prop}[Composition correctness'] Hole-subset simulation is correct w.r.t.\@ composition with automata without holes. \end{prop}
	The idea for the relation on configurations is that the copy of \(A_1\) in \(A\) do the same things as \(A_1\), as if we were trying to prove reflexivity.
	We pose
	\[ R \defobject \mpar{\mpar{s_{1'}, s_2}, s_1} \mapsto \choice{\bigwedge_{v_1 \in V_1} v_{1'} = v_1 & \text{if } s_{1'} = s_1 \\ \bot & \text{otherwise}} \]
		J_2 \cup J_1 \setminus \mbrc{k} \subseteq J_1 \goal{5a} \\
		\sigma_0 \uplus \sigma_{01} \vdash R\mpar{s_0, s_{01}} \goal{5b}
	\end{gather}\vspace{-1cm}
		\forall \mpar{s, s_1} \in S \times S_1, \\
			\bigsymb{\forall} t \defobject \OTg \in \fOT{s}, \bigsymb{\exists} \mpar{t_{1x} \defobject \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X}, \\
			\quad \mpar{\forall x \in X, J'_{1x} \cap J = J' \cap J_1} \\[-10pt]
			\nwedge V \uplus V_1 \uplus \fvars{t} \vdash R\mpar{s, s_1} \wedge g \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
		\end{array}} \\
		\wedge V \uplus V_1 \uplus \biguplus_\subbox{t_1 \in \fOT{s_1}} \fvars{t_1} \vdash R\mpar{s, s_1} \wedge \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \goal{5c}
	\goal{5a} is easy because \(J_2 = \emptyset\).
	For \goal{5b} we use the fact that we don't constrain the behaviour of \(J_2\):
		\iff & \bigwedge_{v_1 \in V_1} \sigma_{01'}\mpar{v_{1'}} = \sigma_{01}\mpar{v_1} \\
	Let \(s \defobject \mpar{s_{1'}, s_2} \in S\) and \(s_1 \in S_1\).

>>>>

% transitions between sections
% TODO: CITER LES CONFERENCES
% TODO: vérifier tous les big([^_]+)_\subbox{[^}]+}[)}]

@@

We begin with Section \ref{sec:notations} by giving some notations used throughout this paper.
Then there are several sections dedicated to define all the interesting objects and properties:
In Section \ref{sec:def} we give a clean and up to date definition of open automata.
In Section \ref{sec:comp} we define their composition without requiring the use of open pNets.
In Section \ref{sec:proofelts} we adapt the standard properties related to refinement relation from LTS to open automata.
After that Section \ref{sec:prelref} is dedicated to progressively build a refinement relation.
Section \ref{sec:refinement} introduces and analyses the refinement relation on open automata.
% future work
% conclusion

@@

The important properties we consider for the relation on states are being a simulation and the prevention of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]\label{def:presim}
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
This definition translated to open automata the classical definition of simulation \marginpar{cite: Fernandez-Mounier-91-b}.
The last formula means that for every pair of related configurations of the automaton and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update. % OR variableS
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \nmm{J \defobject \mpar{\biguplus_\subbox{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}} that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \nmm{J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}} is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

We restrict the holes of the related automata so that we can explain the different refinements without their interactions.
In the following definition, \(\triangle \in \mbrc{\subseteq, =, \supseteq}\).
\begin{defi}[Hole-\(\triangle\) simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\triangle\) simulation of \(A_1\) by \(A_2\) if:
\defitem \(J_1 \triangle J_2\)
\defitem \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\defitem \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
	\mpar{\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} \defobject \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\
		\quad \mpar{\forall x \in X, J'_{2x} \cap J_1 = J'_1 \cap J_2} \\[-10pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap J_1} \beta_{1j} = \beta_{2xj} \\[12pt]
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}
	\end{array}} \\
	\wedge \mpar{V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\end{defi}
The third item of the definition has two parts:
The first part is the main requirement, a (not proved equivalent) reordering of the pre-simulation requirement so that it better fits SMT.
Instead of matching an effective transition to another effective transition as in the pre-simulation definition, we match an open transition to a family of covering open-transitions.
The second part is directly the deadlock prevention requirement.

The hole-equal simulation is applicable when the related automata have the same holes.
It is designed to capture all the refinements other than composition: Duplicating and merging states and transitions, changing variables, removing transitions, adding requirements to guards, specifying more inital variables, and many more.

The hole-subset simulation is applicable when the simulated automaton has less holes.
It is designed to have composition correctness when the composed automata have no holes.

The hole-superset simulation is applicable when the simulated automaton has more holes.
It is designed to capture the addition of holes involved in transitions, which will indirectly add requirements to transitions when the new holes will be composed.
It is also an attempt at giving a meaning to the new holes when composing with any open automata.
However it does not capture the removal of holes because other issues arise when trying to both add and remove holes.

Examples of application of the different refinement relations are given in Appendix \ref{apx:refrel}.
\begin{thm} A hole-\(\triangle\) simulation is a pre-simulation. \end{thm}
Proof is given in Appendix \ref{apx:presim}.
\begin{thm} The hole-\(\triangle\) simulation on automata is a preorder. \end{thm}
% TODO: give intuition, then define R on one line then renamin on the line after
Reflexivity is immediate with \(R \defobject \mpar{s, s'} \mapsto \choice{\bigwedge_{v \in V} v_1 = v_2 & \text{if } s = s'\\ \bot & \text{otherwise}}\) where \(v_x\) is the renaming to distinguish conflicting automaton variables.

For three open automata \(A_1 \defobject \OAg[1]\), \(A_2 \defobject \OAg[2]\) and \(A_3 \defobject \OAg[3]\), such that \(R_{12}\) is a hole-\(\triangle\) simulation of \(A_1\) by \(A_2\) and \(R_{23}\) of \(A_2\) by \(A_3\), we want to define \(R_{13}\) by saying that there exists a state \(s_2 \in S_2\) and values for \(V_2\) such that \(R_{12}\mpar{s_1, s_2} \wedge R_{23}\mpar{s_2, s_3}\).
However we cannot talk about states in formulas.
The solution is to expand the existential quantifier on states as we did in deadlock prevention, which is possible because states are finite.
\nmm{R_{13} \defobject \mpar{s_1, s_3} \mapsto \exists V_2, \bigvee_\subbox{s_2 \in S_2} R_{12}\mpar{s_1, s_2} \wedge R_{23}\mpar{s_2, s_3}} is a witness of the transitivity, proof is given in Appendix \ref{apx:trans}.
\begin{thm}[Context refinement] Hole-equal simulation is context refining. \end{thm}
For three open automata \(A_1 \defobject \OAg[1]\), \(A_2 \defobject \OAg[2]\) and \(A_3 \defobject \OAg[3]\) with \(J_1 = J_2\), let \(k \in J_1\).
Let \(R_{12}\) be a hole-equal simulation of \(A_1\) by \(A_2\) and \(\reach{A\subst{A_3}{k}}\) be a witness that \(A\subst{A_3}{k}\) is non-locking.
The idea for the relation on configurations is that \(A_2\) simulates \(A_1\) while \(A_3\) does the same things in both compositions in the hole \(k\).
\[ R \defobject \mpar{\mpar{s_1, s_{31}}, \mpar{s_2, s_{32}}} \mapsto \choice{R_{12}\mpar{s_1, s_2} \wedge \reach{A_{13}}\mpar{s_1, s_{31}} \wedge \bigwedge_\subbox{v_3 \in V_3} v_{31} = v_{32} & \text{if } s_{31} = s_{32} \\ \bot & \text{otherwise}} \]
Where \(y_{3x}\) is the renaming of \(y_3\) from \(A_3\) in \(A_x\subst{A_3}{k}\).
\begin{prop}[Limited composition correctness]\label{prop:cc'} Hole-subset simulation is correct w.r.t.\@ composition with automata without holes. \end{prop}
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\) with \(J_2 = \emptyset\), let \(k \in J_1\) and \(A \defobject A_1\subst{A_2}{k}\), with \(A = \OAg\).
Let \(\reach{A}\) be a witness that \(A\) is non-locking.
The idea for the relation on configurations is that the copy of \(A_1\) in \(A\) do the same things as \(A_1\), as if we were trying to prove reflexivity.
\[ R \defobject \mpar{\mpar{s_{1'}, s_2}, s_1} \mapsto \choice{\reach{A}\mpar{s_{1'}, s_2} \wedge \bigwedge_\subbox{v_1 \in V_1} v_{1'} = v_1 & \text{if } s_{1'} = s_1 \\ \bot & \text{otherwise}} \]
Where \(y_{1'}\) is the renaming of \(y_1\) from \(A_1\) in \(A\).
\begin{prop}[Special composition correctness]
Let \(A_1 \defobject \OAg[1]\) be hole-superset simulated by \(A_2 \defobject \OAg[2]\).
Let \(j \in J_1 \setminus J_2\) and \(A_3\) an open automaton.
There is a hole-superset simulation of \(A_1\subst{A_3}{j}\) by \(A_2\).
\end{prop}
Proofs of the above theorem and propositions are given in Appendix \ref{apx:ophts}.
